\thetitle

### 2021-06-01 writing/challenges.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

The {\em first challenge\/} stems from the differences between the blame
assignment mechanisms of the three semantic variants.  While Natural assigns
blame to {\em one\/} component, Transient assigns blame to a sequence of
components. The Erasure semantics does not blame components {\it per se\/}, but
it comes with an exception location and a stack trace, which implicitly suggests
fixes.  Each strategy triggers different reactions by the rational programmer
(and real ones, too).

@@

The {\em second challenge\/} is to find a representative collection of programs
with impedance mismatches.\footnote{\citet{cc-oopsla-20} created a collection of
Reticulated Python programs to evaluate their technique of fixing mistakes in
type annotations. Their collection does not come with type-level mistakes in the
code itself.}  The impedance mismatches must represent mistakes that programmers accidentally
create and that the run-time checks of academic systems catch. In other words,
the experiment calls for a collection of mistakes in mix-typed programs that is
representative of those ``in the wild.''  Unfortunately no such collection
exists, and with good reason. The kind of mistakes needed are typically detected
by unit or integration tests; even if it takes some time to find their sources,
these mistakes do not make it into code repositories with appropriate commit
messages.

>>>>

The {\em first challenge\/} stems from the differences between the blame
assignment mechanisms of the three semantic variants.  While Natural assigns
blame to {\em one\/} component, Transient assigns blame to a sequence of
components. The Erasure semantics does not blame components {\it per se\/}, but
it comes with an exception location and a stack trace, which implicitly suggest
fixes.  Each strategy triggers different reactions by the rational programmer
(and real ones, too).

@@

The {\em second challenge\/} is to find a representative collection of programs
with impedance mismatches.\footnote{\citet{cc-oopsla-20} created a collection of
Reticulated Python programs to evaluate their technique of fixing mistakes in
type annotations. Their collection does not come with type-level mistakes in the
code itself.}  The impedance mismatches must represent mistakes that programmers accidentally
create and that the run-time checks of academic systems catch. In other words,
the experiment calls for a collection of mistakes in mixed-typed programs that is
representative of those ``in the wild.''  Unfortunately no such collection
exists, and with good reason. The kind of mistakes needed are typically detected
by unit or integration tests; even if it takes some time to find their sources,
these mistakes do not make it into code repositories with appropriate commit
messages.

### 2021-06-01 writing/discussion.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

Like {\em homo economicus\/}, which idealizes the actual behavior of a
participant in an economy for the sake of mathematical modeling, the model of a
rational programmer idealizes the actual debugging behavior of a software
developer for the sake of a systematic, large-scale analysis. This idealization 
comes with advantages and disadvantages. In the economic realm, mathematical
models have provided some predictive insights into the market's behavior; but as
behavioral economics has shown more recently~\cite{henrich2001search},
the mathematical abstraction of a
rational actor makes predictions also quite unreliable in some situations.
\footnote{It has misled economists to focus on just mathematics, though
this problem is not relevant here---tongue firmly in cheek.}  Just like an
ordinary consumer or producer, an actual software developer is unlikely to stick
to the exact strategy proposed here. When this happens, the predicted benefits
of blame assignment may not materialize. Indeed, the authors' personal
experience suggests such deviations, and it also suggests that deviating often leads to dead-ends.
To make a true judgment of the usefulness of the rational-programmer
idea, the community will need much more experience with this form of evaluation
and relating the evaluation to the behavior of working programmers.

>>>>

Like {\em homo economicus\/}, which idealizes the actual behavior of a
participant in an economy for the sake of mathematical modeling, the model of a
rational programmer idealizes the actual debugging behavior of a software
developer for the sake of a systematic, large-scale analysis. This idealization 
comes with advantages and disadvantages. In the economic realm, mathematical
models have provided some predictive insights into the market's behavior; but as
behavioral economics has shown more recently~\cite{henrich2001search},
the mathematical abstraction of a
rational actor makes predictions also quite unreliable in some situations.\footnote{It
has misled economists to focus on just mathematics, though
this problem is not relevant here---tongue firmly in cheek.}  Just like an
ordinary consumer or producer, an actual software developer is unlikely to stick
to the exact strategy proposed here. When this happens, the predicted benefits
of blame assignment may not materialize. Indeed, the authors' personal
experience suggests such deviations, and it also suggests that deviating often leads to dead-ends.
To make a true judgment of the usefulness of the rational-programmer
idea, the community will need much more experience with this form of evaluation
and relating the evaluation to the behavior of working programmers.

### 2021-06-01 writing/introduction.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

Theoreticians of gradual typing have focused on blame theorems from the
very beginning~\citep{mf-toplas-2009, tf-dls-2006}. ``Well-typed
[components]\footnote{The original authors got this word wrong. A program has many components; blame helps identify a faulty one.} can't be
blamed'' turned the theorem into a slogan~\citep{wf-esop-2009}. Academic
systems (Reticulated Python~\citep{vsc-dls-2019, vss-popl-2017,
vksb-dls-2014} and Typed
Racket~\citep{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010})
come with sophisticated checking and blame assignment strategies (see
sec.~\ref{sec:landscape}). Their academic creators embrace the idea that
blame can help practicing programmers find impedance mismatches, that is,
disagreements between the type ascriptions of a software component and its
behavior.

@@

This paper's {\em first contribution\/} is {\em a method for evaluating the
effectiveness of blame assignment strategies\/} in the gradual typing world.
The top-level innovation is the idea of a {\em rational programmer\/}, that is,
a programmer that acts only in response to available information (see
sec.~\ref{sec:why-rational}). In the case of an impedance mismatch, the
available information consists of the error message and the current state of the
program. The rational programmer can hence use the former to change the
latter---and this systematic, information-driven process can be implemented and
tested, at scale, on real programs.  Turning this idea into a scientific
experiment requires overcoming major challenges: injecting representative
impedance mismatches; putting the various kinds of error information to
comparable use; and sampling the huge space of possibilities (see
sec.~\ref{sec:challenges} and, for details, secs.~\ref{sec:rational} through~\ref{sec:sample}).

The paper's {\it second contribution\/} is a set of {\em results from
applying the evaluation method to three distinct checking regimes for
gradual types and their blame assignment
strategies in approximately 72,000 different scenarios\/} (see
sec.~\ref{sec:results}): (1)~{\it Transient\/}, i.e. Reticulated's
inlined type assertions and collaborative tracking of typed/untyped boundaries~\citep{vss-popl-2017}; (2)~{\it
Natural\/}, i.e. Typed Racket's use of a higher-order contract system and
its blame assignment~\citep{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e.
the approach of industrial systems, which forgo type checks and blame in favor of error
messages from the safety checks of the underlying language. The results
(see sec.~\ref{sec:discussion}) are at least somewhat surprising.  In
principle they validate the conjectures behind the work of theoreticians.
Run-time type checks and blame work together to help with the search for
impedance mismatches between the specified types and the behavior of
untyped components.  Natural's wrapper-based type checks and blame tracking are
more useful than Transient's type in-lined assertions and ``collective blame'' tracking algorithm,
which in turn are superior to Erasure. {\em But\/}, the application of the
method also indicates problems with the expectations of theoreticians. 
In contrast with the theoretical differences between the methods,
{\it Natural\/} is only marginally more useful than {\it Transient\/}, and neither of the two
checking and blame assignment methods are highly superior to {\it Erasure\/}. Additionally, 
the cost of Transient's blame can be huge.  In turn, these
problems suggest that, on one hand, the existing theory does not properly
predict the behavior of blame in real systems with real programs,
and on the other hand, the existing practice lacks data and alternative
experiments to assess the entire landscape of the pragmatics of blame.

>>>>

Theoreticians of gradual typing have focused on blame theorems from the
very beginning~\citep{mf-toplas-2009, tf-dls-2006}. ``Well-typed
[components]\footnote{The original authors got this word wrong. A program has many components; blame helps identify a faulty one.} can't be
blamed'' turned the theorem into a slogan~\citep{wf-esop-2009}. Academic
systems (Reticulated Python~\citep{vsc-dls-2019, vss-popl-2017,
vksb-dls-2014} and Typed
Racket~\citep{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010})
come with sophisticated checking and blame assignment strategies
(sec.~\ref{sec:landscape}). Their academic creators embrace the idea that
blame can help practicing programmers find impedance mismatches, that is,
disagreements between the type ascriptions of a software component and its
behavior.

@@

This paper's {\em first contribution\/} is {\em a method for evaluating the
effectiveness of blame assignment strategies\/} in the gradual typing world.
The top-level innovation is the idea of a {\em rational programmer\/}, that is,
a programmer that acts only in response to available information
(sec.~\ref{sec:why-rational}). In the case of an impedance mismatch, the
available information consists of the error message and the current state of the
program. The rational programmer can hence use the former to change the
latter---and this systematic, information-driven process can be implemented and
tested, at scale, on real programs.  Turning this idea into a scientific
experiment requires overcoming major challenges: injecting representative
impedance mismatches; putting the various kinds of error information to
comparable use; and sampling the huge space of possibilities (see
sec.~\ref{sec:challenges} and, for details, secs.~\ref{sec:rational} through~\ref{sec:sample}).

The paper's {\it second contribution\/} is a set of {\em results from
applying the evaluation method to three distinct checking regimes for
gradual types and their blame assignment
strategies in approximately 72,000 different scenarios\/}
(sec.~\ref{sec:results}): (1)~{\it Transient\/}, i.e. Reticulated's
inlined type assertions and collaborative tracking of typed/untyped boundaries~\citep{vss-popl-2017}; (2)~{\it
Natural\/}, i.e. Typed Racket's use of a higher-order contract system and
its blame assignment~\citep{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e.
the approach of industrial systems, which forgo type checks and blame in favor of error
messages from the safety checks of the underlying language. The results
(sec.~\ref{sec:discussion}) are at least somewhat surprising.  In
principle they validate the conjectures behind the work of theoreticians.
Run-time type checks and blame work together to help with the search for
impedance mismatches between the specified types and the behavior of
untyped components.  Natural's wrapper-based type checks and blame tracking are
more useful than Transient's type in-lined assertions and ``collective blame'' tracking algorithm,
which in turn are superior to Erasure. {\em But\/}, the application of the
method also indicates problems with the expectations of theoreticians. 
In contrast with the theoretical differences between the methods,
{\it Natural\/} is only marginally more useful than {\it Transient\/}, and neither of the two
checking and blame assignment methods are highly superior to {\it Erasure\/}. Additionally, 
the cost of Transient's blame can be huge.  In turn, these
problems suggest that, on one hand, the existing theory does not properly
predict the behavior of blame in real systems with real programs,
and on the other hand, the existing practice lacks data and alternative
experiments to assess the entire landscape of the pragmatics of blame.

### 2021-06-01 writing/landscape.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

As this analysis implies, if a return-type check fails, the problem is
that the untyped module supplied a function, here {\tt
pack-lib}, that is not a  match for the type ascribed by the typed module. 
Hence either the type at the boundary between the two modules is wrong or, 
if the programmer trusts the type, the untyped module is at fault.
If an argument-type check fails, responsibility  lies with
{\tt client}. After all, either the type it ascribes to the argument is
wrong or the argument it produced clashes with the type. Due to 
proxies, {\it Natural\/} can easily track the 
boundary, type and responsible parties that correspond to each check. 
Thus, in the example of figure~\ref{fig:example}, as
\texttt{pack} returns, the return-type check fails and Natural blames
the boundary between \texttt{pack-lib} with {\tt typed-pack-lib} and
{\tt crypto-pack-lib}, respectively, for the two {\tt define}s in {\tt client}.

@@

In order to locate the corresponding boundaries for failed checks, Transient
maintains a map from values to the boundaries between typed and untyped modules that
they cross, plus the corresponding types. In the example, the map records that
\texttt{pack} crosses from \texttt{pack-lib} to {\tt typed-pack-lib} and
from {\tt pack-lib} to {\tt crypto-pack-lib} with the type that appears in the
{\tt required/typed} forms in the example. Since the failed check corresponds to
the return type of \texttt{pack}, assuming that the type is correct, 
the responsible parties are the sources of the two
boundary crossings, \texttt{pack-lib}. In
general though, Transient blames more than one boundary. In fact, the theoretical
work of~\citet{gfd-oopsla-2019} shows that for some programs Transient
constructs a blame sequence that excludes responsible parties and includes modules
irrelevant to the failing check.

>>>>

As this analysis implies, if a return-type check fails, the problem is
that the untyped module, here {\tt pack-lib}, supplied a function that is not a
match for the type ascribed by the typed module.
Hence either the type at the boundary between the two modules is wrong or, 
if the programmer trusts the type, the untyped module is at fault.
If an argument-type check fails, responsibility  lies with
{\tt client}. After all, either the type it ascribes to the argument is
wrong or the argument it produced clashes with the type. Due to 
proxies, {\it Natural\/} can easily track the 
boundary, type, and responsible parties that correspond to each check. 
Thus, in the example of figure~\ref{fig:example}, as
\texttt{pack} returns, the return-type check fails and Natural blames
the boundary between \texttt{pack-lib} with {\tt typed-pack-lib} and
{\tt crypto-pack-lib}, respectively, for the two {\tt define}s in {\tt client}.

@@

In order to locate the corresponding boundaries for failed checks, Transient
maintains a map from values to the boundaries between typed and untyped modules that
they cross, plus the corresponding types. In the example, the map records that
\texttt{pack} crosses from \texttt{pack-lib} to {\tt typed-pack-lib} and
from {\tt pack-lib} to {\tt crypto-pack-lib} with the type that appears in the
{\tt required/typed} forms in the example. Since the failed check corresponds to
the return type of \texttt{pack}, assuming that the type is correct,
the responsible party is the source of the two
boundary crossings: \texttt{pack-lib}. In
general though, Transient blames more than one boundary. In fact, the theoretical
work of~\citet{gfd-oopsla-2019} shows that for some programs Transient
constructs a blame sequence that excludes responsible parties and includes modules
irrelevant to the failing check.

### 2021-06-01 writing/mutate-benchmarks.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

\citet{gtnffvf-jfp-2019}'s collection of Typed Racket programs for
systematically measuring the implementation's performance satisfies these
criteria. The benchmark suite consists of fully typed, correct programs, written
by a number of different authors, who have maintained and evolved these programs
over several years. The programs range widely in size, complexity, purpose,
origin and in programming style. They rely on many Typed Racket features:
occurrence typing~\citep{tf-icfp-2010}, types for mutable and immutable data
structures~\citep{hpst-sfp-2010}, types for first-class classes and
objects~\citep{tsdtf-oopsla-2012}, and types for Racket's numeric
tower~\citep{stathff-padl-12}.  Finally, all of the programs are deterministic,
so any changes in the programs' behavior between runs can be solely attributed
to the actions of the rational programmer.

Figure~\ref{table:benchmark-descriptions} describes the ten benchmark programs that
meet all the criteria, and furthermore come with the largest dependency graphs.
This additional filter reflects that Typed Racket demands type assignments
for entire modules, and so finding errors in benchmarks with small dependency graphs
would almost be trivial.

>>>>

\citet{gtnffvf-jfp-2019}'s collection of Typed Racket programs for
systematically measuring the implementation's performance satisfies these
criteria. The benchmark suite consists of fully typed, correct programs, written
by a number of different authors who had maintained and evolved these programs
over time. The programs range widely in size, complexity, purpose,
origin and in programming style. They rely on many Typed Racket features:
occurrence typing~\citep{tf-icfp-2010}, types for mutable and immutable data
structures~\citep{hpst-sfp-2010}, types for first-class classes and
objects~\citep{tsdtf-oopsla-2012}, and types for Racket's numeric
tower~\citep{stathff-padl-12}.  Finally, all of the programs are deterministic,
so any changes in the programs' behavior between runs can be solely attributed
to the actions of the rational programmer.

Figure~\ref{table:benchmark-descriptions} describes the ten benchmark programs that
meet all the criteria, and furthermore come with the largest dependency graphs.
This additional filter reflects that Typed Racket demands type assignments
for entire modules, and so finding errors in benchmarks with small dependency graphs
would almost be trivial for the rational programmer.

### 2021-06-01 writing/mutate.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

Putting the rational programmer to work means generating many mutants and turning
those into debugging scenarios. The process must start with a suitable
collection of representative programs (see sec.~\ref{sub:mutate-benchmarks}).
Since existing mutators do not generate useful mutants, the next step is to
develop new mutators (see sec.~\ref{sub:mutate-mutators}) and to validate their
suitability on the benchmarks (see sec.~\ref{sub:mutate-interesting}).

>>>>

Putting the rational programmer to work means generating many mutants and turning
those into debugging scenarios. The process must start with a suitable
collection of representative programs (sec.~\ref{sub:mutate-benchmarks}).
Since existing mutators do not generate useful mutants, the next step is to
develop new mutators (sec.~\ref{sub:mutate-mutators}) and to validate their
suitability on the benchmarks (sec.~\ref{sub:mutate-interesting}).

### 2021-06-01 writing/rational-stuff.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

The actions of the rational programmer create an ascending chain---dubbed a
\emph{trail}---in $\lattice{\system}$.  The {\em root\/} $\conf_0$ of a trail is
the initial debugging scenario. If the program for some scenario  $\conf_i$ type-checks,
the rational programmer runs the program until it raises a run-time error.  The
rational programmer then uses the information in the error to decide which module to
equip with types. This choice constructs scenario $\conf_{i+1}$ from $\conf_{i}$
and thus creates a trail. The trail's construction ends when it reaches a
scenario that contains the mutated module because the type checker rejects its
typed version outright.  At this point, the source of the impedance mismatch is
identified.

>>>>

The actions of the rational programmer create an ascending chain---dubbed a
\emph{trail}---in $\lattice{\system}$.  The {\em root\/} $\conf_0$ of a trail is
the initial debugging scenario. If the program for some scenario  $\conf_i$ type-checks,
the rational programmer runs the program until it raises a run-time error.  The
rational programmer then uses the information in the error to decide which module to
equip with types. This choice constructs scenario $\conf_{i+1}$ from $\conf_{i}$
and thus lengthens the trail. A trail's construction ends successfully when it reaches a
scenario that contains the mutated module because the type checker rejects its
typed version outright.  At this point, the source of the impedance mismatch is
identified.

### 2021-06-01 writing/rational-transient.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

The Transient semantics assigns blame to a sequence of modules. The blame
assignment says that the value witnessing the impedance mismatch may have
crossed the boundaries between elements in the sequence, and each crossing
checked the value's type in a shallow manner.

This ambiguity in Transient blame raises the question of how the rational
programmer should react when the language produces a blame set. Our answer
is that the rational programmer has at least two reasonable options. The first
one is to select the untyped module that is added to the blame sequence first and
assign types to only that one---after all, if fully checked, the types of this
first module should be able to detect an impedance mismatch earlier in the
evaluation of a program than the later ones. The second option is to select the
module that is added to the blame sequence last, effectively interpreting the
blame sequence as a boundary-aware stack.

>>>>

The Transient semantics assigns blame to a sequence of modules. The blame
assignment says that the value witnessing the impedance mismatch may have
crossed the boundaries between elements in the sequence, and that each crossing
checked the value's type in a shallow manner.

This ambiguity in Transient blame raises the question of how the rational
programmer should react when the language produces a blame sequence. Our answer
is that the rational programmer has at least two reasonable options. The first
one is to select the untyped module that is added to the blame sequence first and
assign types to only that one---after all, if fully checked, the types of this
first module should be able to detect an impedance mismatch earlier in the
evaluation of a program than the later ones. The second option is to select the
module that is added to the blame sequence last, effectively interpreting the
blame sequence as a boundary-aware stack.

### 2021-06-01 writing/rational.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

Section~\ref{sec:why-rational} explains how a migratory type setting helps with
finding the source of an impedance mismatch. Roughly speaking, it encourages the
rational programmer to equip a module with types if it is blamed in an error
message. A sequence of such steps makes up a path in the
lattice of type migration~\citep{tfgnvf-popl-2016}. The lattice describes the
space in which the modes of the rational programmer search for bugs (see
sec.~\ref{sub:stuff}).
%\mf{what does this sentence really mean? what does it contribute?}
%\ll{Addressed}

Each mode receives different kinds of information and thus may construct
different paths in the lattice. As discussed in section~\ref{sec:challenges},
evaluating blame relies on comparing modes of the rational programmer
within the same semantics and across different semantics.  
Hence the research problem is to develop modes for the rational programmer and to make
them comparable even when they correspond to different semantics and process
different kinds of information (see secs.~\ref{sub:natural}
through~\ref{sub:erasure}). Programmer effort relative to a fixed debugging
scenario introduces another dimension along which the modes become may be
compared (see sec.~\ref{sub:effort}). With these notions in place, it
becomes possible to state the experimental questions and describes the process (see sec.~\ref{sub:experiment}).

>>>>

Section~\ref{sec:why-rational} explains how a migratory type setting helps with
finding the source of an impedance mismatch. Roughly speaking, it encourages the
rational programmer to equip a module with types if it is blamed in an error
message. A sequence of such steps makes up a path in the
lattice of type migration~\citep{tfgnvf-popl-2016}. The lattice describes the
space in which the modes of the rational programmer search for bugs
(sec.~\ref{sub:stuff}).
%\mf{what does this sentence really mean? what does it contribute?}
%\ll{Addressed}

Each mode receives different kinds of information and thus may construct
different paths in the lattice. As discussed in section~\ref{sec:challenges},
evaluating blame relies on comparing modes of the rational programmer
within the same semantics and across different semantics.  
Hence the research problem is to develop modes for the rational programmer and to make
them comparable even when they correspond to different semantics and process
different kinds of information (secs.~\ref{sub:natural}
through~\ref{sub:erasure}). Programmer effort relative to a fixed debugging
scenario introduces another dimension along which the modes become may be
compared (sec.~\ref{sub:effort}). With these notions in place, it
becomes possible to state the experimental questions and describes the process (sec.~\ref{sub:experiment}).

### 2021-06-01 writing/related.tex
fixing prose
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks p

%% -----------------------------------------------------------------------------

At a philosophical level, \citet{lksfd-popl-2020} present the first empirical
analysis {\em without\/} involving humans in this general area.  While they do
not spell out the notion of the rational programmer, they present many of the
basic ingredients.

At a technical level, the two pieces of work differ in many ways.  First the
experiment presented here involves three different modes of gradual typing,
theirs a single notion of contract checking.  Second, this paper also
contributes the idea of creating three comparable rational programmers, with
several modes. Hence it can answer whether blame adds value to a gradual type
system and which gradual type system it benefits the most, while theirs
establishes only a systematic relation between blame for contracts and
behavioral bugs. Finally, this paper also contributes type-level mutators. As
explained in section~\ref{sec:mutate}, almost none of~\citet{lksfd-popl-2020}'s
mutators is useful in the context of gradual typing.

The literature on gradual typing presents many semantics beyond the three used
here and three additional blame strategies.  Pyret (pyret.org) assigns fixed-size
data types the Natural semantics and functions a Transient semantics. The
Forgetful~\citep{cl-icfp-2017} and the Amnesic~\citep{gfd-oopsla-2019} semantics
are similar to Transient but use wrappers instead of in-lined checks.
Nom~\citep{mt-oopsla-2017} and other \emph{concrete\/}
semantics~\citep{wnlov-popl-2010, rsfbv-popl-2015, rzv-ecoop-2015,
rat-oopsla-2017} assume that every value comes with a type tag and use tag
checks to supervise the interactions between typed and untyped code.  The semantics~\citep{tlt-popl-2019,
etg-icfp-19, tt-scp-20, tgt-popl-18, tt-sas-17} derived with the Abstracting
Gradual Typing technique~\citep{gct-popl-2016} are variants of Natural.
The Monotonic semantics~\citep{svctg-esop-2015, rsfbv-popl-2015, sfrbcsb-popl-2014, kas-pldi-2019} 
differs from Natural in the treatment of mutable data. It 
associates every heap location with a type and permits only updates that 
do not make the type less precise.  
Among these semantics, only Amnesic, Nom, and Monotonic present interesting blame
strategies.  The experiment excludes the first because it is 
merely a theoretical construction, the second because it imposes severe restrictions on
programmers, and the third because it requires a re-engineering of the Racket runtime.

>>>>

\newcommand{\pyret}{\href{https://www.pyret.org}{pyret.org}}
%% -----------------------------------------------------------------------------

At a philosophical level, \citet{lksfd-popl-2020} present the first empirical
analysis {\em without\/} involving humans in this general area.  While they do
not spell out the notion of the rational programmer, they present many of the
basic ingredients. At a technical level, the two pieces of work differ in many
ways.  First the experiment presented here involves three different modes of
gradual typing, theirs a single notion of contract checking.  Second, this paper
also contributes the idea of creating three comparable rational programmers,
with several modes. Hence it can answer whether blame adds value to a gradual
type system and which gradual type system it benefits the most, while theirs
establishes only a systematic relation between blame for contracts and
behavioral bugs. Finally, this paper also contributes type-level mutators. As
explained in section~\ref{sec:mutate}, almost none of~\citet{lksfd-popl-2020}'s
mutators are useful in the context of gradual typing.

The literature on gradual typing presents many semantics beyond the three used
here, and three additional blame strategies.  Pyret (\pyret{}) assigns fixed-size
data types the Natural semantics and functions a Transient semantics. The
Forgetful~\citep{cl-icfp-2017} and the Amnesic~\citep{gfd-oopsla-2019} semantics
are similar to Transient but use wrappers instead of in-lined checks.
Nom~\citep{mt-oopsla-2017} and other \emph{concrete\/}
semantics~\citep{wnlov-popl-2010, rsfbv-popl-2015, rzv-ecoop-2015,
rat-oopsla-2017} assume that every value comes with a type tag and use tag
checks to supervise the interactions between typed and untyped code.  The semantics derived with the Abstracting
Gradual Typing technique~\citep{gct-popl-2016} are variants of Natural.
The Monotonic semantics~\citep{svctg-esop-2015, rsfbv-popl-2015, sfrbcsb-popl-2014, kas-pldi-2019} 
differs from Natural in the treatment of mutable data. It 
associates every heap location with a type and rejects updates that lower the
precision of types.
Among these semantics, only Amnesic, Nom, and Monotonic present interesting blame
strategies.  The experiment excludes the first because it is 
merely a theoretical construction, the second because it imposes severe restrictions on
programmers, and the third because it requires a re-engineering of the Racket runtime.

### 2021-06-01 writing/why-rational-methodology.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

\item The error messages of blame-assignment mechanisms include suspect
locations at the boundary of typed and untyped code fragments.
The Wadler-Findler slogan suggests that the source of the problem is concealed due to a lack of types, so
adding types to the untyped fragment should lead to the source of the impendance mismatch.

>>>>

\item The error messages of blame-assignment mechanisms include suspect
locations at the boundary of typed and untyped code fragments.
The Wadler--Findler slogan suggests that the source of the problem is concealed due to a lack of types, so
adding types to the untyped fragment should lead to the source of the impendance mismatch.
